Finiteness in Cubical Type Theory

### 2020-01-14 paper.tex
finite isomorphisms

  We formalise our work in Cubical Agda, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
\subsection{Strong Finiteness Predicates}
\input{figures/finite-classification}
We will first explore the ``strong'' notions of finiteness (i.e. those at least
as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
with a special focus on cardinal finiteness (section~\ref{cardinal}), and
manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
our knowledge.

>>>>

  We formalise our work in Cubical Agda, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
In constructive mathematics, we are often preoccupied with \emph{how} we know
something.
Finiteness is one of the prime examples which demonstrates this phenomenon: to
show that some type is finite, we might provide a bijection with another
finite set, or a surjection from some finite type.
We could alternatively say that a type is finite if any collection of its
elements larger than some bound contains duplicates, or if any stream of its
elements contains duplicates.

Classically, all of these ways to prove finiteness end up proving the same
thing: constructively, however, each defines a fundamentally different
``version'' of finiteness.
In Martin-Löf type theory \cite{martin-lofIntuitionisticTypeTheory1980}, these
variations (and more) have been explored in
\cite{spiwackConstructivelyFinite2010} and
\cite{firsovVariationsNoetherianness2016}.

Homotopy Type Theory \cite{hottbook} further adds to this story in two ways.
First, its universe of types includes non-sets: whether or not a finiteness
predicate includes these gives us another dimension on which to discriminate.
Secondly,

classifying these types as
finite or not adds

can step in.
Through primarily a generalisation of equality, we gain access to a number of
exotic types which can, among other things, allow us to perform this ``hiding''.
In the context of finiteness, the act of hiding is worth studying in its own
right, and further delineates several new finiteness predicates which collapse
even in standard

The other headline feature of HoTT's generalised equality is univalence: put
simply, it allows us to treat isomorphic types as equal, giving us in the
constructive world access to a technique that is commonplace classically.
It turns out that finite types have a lot to say about isomorphisms, and we will
rely heavily on this aspect of HoTT to work with them.
\subsection{Contributions}
In this work, we will explore finite types in Cubical Type Theory
\cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
types, species, and demonstrate their practical uses for proofs in dependently
typed programming languages.

In section~\ref{strong-finiteness}, we will explore the ``strong'' notions of
finiteness (i.e. those at least as strong as Kuratowski finiteness
\cite{kuratowskiNotionEnsembleFini1920}), with a special focus on cardinal
finiteness.
We will delineate some subtly different finiteness predicates, and prove the
relationship between each.
We will prove, then, that every discrete Kuratowski finite type is cardinal
finite, and subsequently will show that discrete Kuratowski finite sets form a
topos.

In section~\ref{species-and-containers}, we will redefine our finiteness
predicates using \emph{containers}, and prove that the new definitions are
equivalent to the old.
Using this, we will prove much stronger relationships between our finiteness
predicates and relation-based definitions of finiteness.
Finally, We will show how the species-container duality falls out naturally from
these new definitions.

In section~\ref{infinite-cardinalities}, we will extend our study of finite
types to infinite but countable types.
We will see that the finiteness predicates are mirrored with countable
counterparts, and we will prove closure under the Kleene star and plus.

Proofs of finiteness have well-known practical applications in
constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
In section~\ref{practical}, we build a library which exploits these uses in
Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
complex proofs over finite types.
We frame this in terms of the principle of omniscience for finite types.
Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
go further than the usual examples of this kind of proof automation: as well as
proving properties about functions, we can synthesise functions whole-cloth from
their desired properties.
Through the unified interface for finite and countable types, we can
reuse the automation machinery for \emph{partial} proof search over infinite
search spaces.
Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
prove isomorphisms between the curried and uncurried forms of \(n\)-ary
dependent functions.

% Classically, of course, each of these statements are equivalent.
% Constructively, though, all four predicates are fundamentally different.

% Part of the difference comes from ``leaking'' information about the finite type
% in the proof of its finiteness.
% If I demonstrate a type is finite by showing a surjection to that type, I've
% revealed a way to construct all of the elements of my type.
% If I show that it is bounded, I reveal an upper bound on the cardinality of the
% type.

\subsection{Strong Finiteness Predicates}
\input{figures/finite-classification}
We will first explore the ``strong'' notions of finiteness (i.e. those at least
as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
with a special focus on cardinal finiteness (section~\ref{cardinal}), and
manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
our knowledge.
